---
title: "Exercise 8"
author: "Jonas Freimuth, 807218"
date: "14 12 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("lme4")
library("dplyr")
```

## Task 1

Narrow sense heritability: 

$$h^2 = \frac{\sigma_A^2}{\sigma_P^2}$$

$$Cov(FullSibs) = \frac{\sigma_A^2}{2} + \frac{\sigma_D^2}{4},$$ 
here: $\sigma_D^2 = 0$

$$\Rightarrow \sigma_A^2 = 2 * Cov(FullSibs) = 2 * \sigma_{family}^2$$

```{r task_1_1}
biom_dat <- read.csv("data/Ex8_1.csv")

# this method of extracting family variance was inspired by Philipps solution
biom_lme <- lmer(biomass ~ (1 | family), biom_dat)

vc <- VarCorr(biom_lme)

var_fam  <- as.numeric(vc[1])

var_phen <- var(biom_dat$biomass)

h2_biom  <- (2 * var_fam) / var_phen
```

$\Rightarrow$ The narrow sense heritablility of biomass $h_{biomass}^2$
is $`r round(h2_biom, 2)`$

## Task 2

```{r task_2_1}
yld_dat <- read.csv("data/Ex8_2.csv")

yld_dat_lme_nest <- lmer(yield ~ (1 | sire/dame), yld_dat)

vc <- VarCorr(yld_dat_lme_nest)

var_dame <- as.numeric(vc[1])
var_sire <- as.numeric(vc[2])

# see Exercise 6: Task 3 ii
sig_A <- 4 *  var_sire
sig_D <- 4 * (var_sire + var_dame) - 2 * (sig_A)

var_phen <- var(yld_dat$yield)

h2_yld <- sig_A / var_phen
```

$\Rightarrow$ The narrow sense heritablility of the yield $h_{yield}^2$ is
$`r round(h2_yld, 2)`$

## Task 3

```{r task_3_1}
bp_1_dat <- read.csv("data/Ex8_3.csv")

cov_po <- with(bp_1_dat, cov(offspring, parent))
var_p  <- var(bp_1_dat$parent)

h2_bp_1 <- cov_po / var_p
```

$\Rightarrow$ The narrow sense heritablility of the blood pressure under the 
male offspring design $h_{blood\ pressure, 1}^2$ is $`r round(h2_bp_1, 2)`$


## Task 4
```{r task_4_1}
bp_2_dat <- read.csv("data/Ex8_4.csv") %>% 
  rowwise() %>% 
  mutate(mean_parent = mean(parent1, parent2))

cov_mpo <- with(bp_2_dat, cov(offspring, mean_parent))
var_mp  <- var(bp_2_dat$mean_parent)

h2_bp_2 <- cov_mpo / var_mp
```

$\Rightarrow$ The narrow sense heritablility of the blood pressure under the 
two parent design $h_{blood\ pressure, 2}^2$ is $`r round(h2_bp_2, 2)`$
