---
title: "Exercise 3"
author: "Jonas Freimuth, 807218"
date: "12 11 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require("lme4",       quietly = TRUE)
require("lmerTest",   quietly = TRUE)
require("dplyr",      quietly = TRUE)
require("tidyr",      quietly = TRUE)
require("tidyselect", quietly = TRUE)
require("ggplot2",    quietly = TRUE)

# disable grouping output message
options(dplyr.summarise.inform = FALSE)

```

## Task 1

```{r task_1_1}
wh_df <- read.csv("data/Ex3.csv") %>% 
  mutate(sire = as.factor(sire),
         dame = as.factor(dame))

str(wh_df)
```

```{r task_1_2}
ggplot(wh_df, aes(soil, yield, col = sire, group = sire)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  theme_minimal()
```

$\Rightarrow$ As the regression lines are (almost) parallel, different sires 
lead to the same slope but different intercepts.

```{r task_1_3}
ggplot(wh_df, aes(soil, yield, col = dame, group = dame)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  theme_minimal() + 
  theme(legend.position = "none")
```

$\Rightarrow$ Both intercepts and slopes vary between dame groups.

## Task 2

```{r task_2_1}
mod_ranint_sum <- summary(lmer(yield ~ soil + (1 | sire), wh_df))
mod_ranint_sum

ranint_coef <- round(coef(mod_ranint_sum), 3)
```

### Fixed Effects

Intercept: $\beta_{Incp} = `r ranint_coef[1, 1]`, p = `r ranint_coef[1, 5]`$  
Eff. of Soil: $\beta_{soil} = `r ranint_coef[2, 1]`,$
$p = `r ranint_coef[2, 5]`$  

### Random Effects

Variance of Sire Intercept: $Var_{Incp, sire} = 0.885$  
Residual variance: $Var_{Resid} = 1.421$

## Task 3

```{r task_3_1}
mod_ranslp_sum <- summary(lmer(yield ~ soil + (0 + soil | dame), wh_df))
mod_ranslp_sum

ranslp_coef <- round(coef(mod_ranslp_sum), 3)
```

### Fixed Effects

Intercept: $\beta_{Incp} = `r ranslp_coef[1, 1]`, p = `r ranslp_coef[1, 5]`$  
Eff. of Soil: $\beta_{soil} = `r ranslp_coef[2, 1]`,$
$p = `r ranslp_coef[2, 5]`$  

### Random Effects

Variance of dame effect on yield: $Var_{slope, dame} = 0.0176$  
Residual variance: $Var_{Resid} = 1.372$

## Task 4

For the sire effect only a random intercept will be used, while the dame effect 
will be modeled using a random intercept and a random slope with correlation.

```{r task_4_1}
mod_raneff_sum <- summary(lmer(yield ~ soil + (1 | sire) + (soil | dame),
                               wh_df))
mod_raneff_sum

raneff_coef <- round(coef(mod_raneff_sum), 3)
```

### Fixed Effects

Intercept: $\beta_{Incp} = `r raneff_coef[1, 1]`, p = `r raneff_coef[1, 5]`$  
Eff. of Soil: $\beta_{soil} = `r raneff_coef[2, 1]`,$
$p = `r raneff_coef[2, 5]`$  

### Random Effects

Variance of dame intercept: $Var_{Incp, sire} = 1.861$  
Variance of dame effect on yield: $Var_{slope, dame} = 0.017$  
Variance of sire intercept: $Var_{Incp, sire} = 0.719$  
Residual variance: $Var_{Resid} = 0.930$
