---
title: "Exercise 6"
author: "Jonas Freimuth, 807218"
date: "1 12 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("lme4")
library("lmerTest")
```

## Task 1

```{r task_1_1}
mz_yld_1 <- read.csv("data/Ex5_1.csv")

mz_yld_lme <- lmer(yield ~ (1 | geno) + (1 | env) + (1 | geno:env), mz_yld_1)
summary(mz_yld_lme)
```

Linear Mixed Effects Model variance components:  
* Genetic:      3.789  
* Environment: 18.535  
* Interaction:  0.648  

```{r task_1_2}
rnd_yld_aov <- EMSanova(yield ~ geno + env, mz_yld_1, type = c("R", "R"))

n <- 6
a <- 3
b <- 2

gen_MS <- rnd_yld_aov$MS[1]
env_MS <- rnd_yld_aov$MS[2]
int_MS <- rnd_yld_aov$MS[3]
err_MS <- rnd_yld_aov$MS[4]

gen_var_com <- (gen_MS - int_MS) / (b * n) 
env_var_com <- (env_MS - int_MS) / (a * n) 
int_var_com <- (int_MS - err_MS) /   (n)   

cat("ANOVA Variance components:\n")
cat("\tGenetic:\t",     gen_var_com, "\n")
cat("\tEnvironment:\t", env_var_com, "\n")
cat("\tInteraction:\t", int_var_com, "\n")
```

$\Rightarrow$ Both ways of determine variance components give the same result.

## Task 2

### i)
```{r task_2_1}
ranef(mz_yld_lme)$geno
```


### ii)

```{r task_2_2}
mz_yld_lme_alt <- lmer(yield ~ (1 | geno) + (1 | env), mz_yld_1)

anova(mz_yld_lme, mz_yld_lme_alt, test = "LRT")
```

$\Rightarrow$ The interaction effect of genotype and environmtent is significant.  

## Task 3

### i)

```{r task_3_1}
mz_yld_2 <- read.csv("data/Ex5_2.csv")

mz_yld_lme_nst <- lmer(yield ~ (1 | sire/dame) + (1 | sire), mz_yld_2)
summary(mz_yld_lme_nst)
```



```{r task_3_2}
nst_yld_aov <- EMSanova(yield ~ sire + dame, mz_yld_2, type = c("R", "R"),
                        nested = c(NA, "sire"))

n <- 21
a <- 2
b <- 5

sir_MS <- nst_yld_aov$MS[1]
dam_MS <- nst_yld_aov$MS[2]
err_MS <- nst_yld_aov$MS[3]

sir_var_com <- (sir_MS - dam_MS) / (n * b)
dam_var_com <- (dam_MS - err_MS) /   (n)

cat("ANOVA Variance components:\n")
cat("\tSire:\t\t\t",            sir_var_com, "\n")
cat("\tDame nested in Sire:\t", dam_var_com, "\n")
```

$\Rightarrow$ As the sum of the sire effects above is $1.436$, the ANOVA also
agrees with the LME model here.


### ii)

Those accessions sharing both parents are full-sib (e.g. all combinations of 
sire 1 and dame 1), while those that share only one parent are half-sib (e.g. 
combinations of sire 1 with dame 1 to 5).

As $Cov(HalfSibs) = \frac{\sigma_A^2}{4}$ and 
$Cov(FullSibs) = \frac{\sigma_A^2}{2} + \frac{\sigma_D^2}{4}$ $\Rightarrow$ 
$\sigma_A^2 = 4 * Cov(HalfSibs)$ and $\sigma_D^2 = 4 * Cov(FullSibs) - 2 \sigma_A^2$

```{r task 3_3}
cov()
```

