---
title: "Exercise 2"
author: "Jonas Freimuth, 807218"
date: "3 11 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require("dplyr",      quietly = TRUE)
require("tidyr",      quietly = TRUE)
require("tidyselect", quietly = TRUE)
```

## Task 1

```{r task_1_1}
aov_tab <- data.frame("Sum Sq"  = c(720, 480, 1200),
                      "df"      = c(3, 16, 19),
                      "Mean Sq" = c(240, 30, NA),
                      "F val"   = c(8, NA, NA),
                      row.names = c("Treatment", "Error", "Total"))

aov_tab
```

Total degrees of freedom: $N - 1 = 19 \Rightarrow N = 20 \Rightarrow$ $20$ 
samples overall. Treatment degrees of freedom: $m - 1 = 3 \Rightarrow m = 4$ 
$\Rightarrow$ four treatment groups. Samples per group assuming perfect balance
of groups: $n = N / m = 20 / 4 = 5 \Rightarrow$ five samples per group.

There is a significant difference between the means of different treatments 
(ANOVA, $F_{3, 16} = 8, p = `r round(pf(8, 3, 16, lower.tail = FALSE), 3)`$).

## Task 2

```{r task_2_1}
mz_yld_raw <- read.csv("data/Ex2.csv")

str(mz_yld_raw)

mz_yld <- mz_yld_raw %>% 
  select(matches("E\\d_F\\d")) %>% 
  pivot_longer(cols      = everything(),
               names_to  = c("env", "fam"),
               names_sep = "_",
               values_to = "yield")
  

head(mz_yld)
```

```{r task_2_test}
mod_yld_f1 <- aov(yield ~ env, data = mz_yld %>% filter(fam == "F1"))

summary(mod_yld_f1)
```


```{r task_2_2}
mz_yld_f1 <- mz_yld %>% 
  filter(fam == "F1")

m <- length(unique(mz_yld_f1$env))
n <- length(mz_yld_f1$env[mz_yld_f1$env == unique(mz_yld_f1$env[1])])
N <- nrow(mz_yld_f1)

# calculate degrees of freedom
dfs <- c("betw" = m - 1,
         "ins"  = N - m,
         "tot"  = N - 1)

# calculate means
grp_means <- mz_yld_f1 %>% 
  group_by(env) %>% 
  summarise(mean = mean(yield))

gm <- mean(mz_yld_f1$yield)

# calculate sums of squares
ins_sum_sq_vec <- rep(NA, length(mz_yld_f1$yield))

for (i in 1:m) {
  
  ins_sum_sq <- mz_yld_f1$yield[mz_yld_f1$env == unique(mz_yld_f1$env)[i]]
  ins_sum_sq <- (ins_sum_sq - grp_means$mean[i]) ^ 2
  ins_sum_sq_vec[(((i - 1) * n) + 1):((i * n))] <- ins_sum_sq
  
}

sum_sq <- c("betw" = sum((grp_means$mean  - gm) ^ 2 * n),
            "ins"  = sum(ins_sum_sq_vec),
            "tot"  = sum((mz_yld_f1$yield - gm) ^ 2))

# calculate mean squares
mean_sq <- c("betw" = sum_sq[["betw"]] / dfs[["betw"]],
             "ins"  = sum_sq[["ins" ]] / dfs[["ins" ]],
             "tot"  = NA)

# calculate F value
f_val <- c("betw" = mean_sq[["betw"]] / mean_sq[["ins"]],
           "ins"  = NA,
           "tot"  = NA)

# assemble table
aov_tab <- data.frame("Sum Sq"  = sum_sq,
                      "df"      = dfs,
                      "Mean Sq" = mean_sq,
                      "F val"   = f_val,
                      row.names = c("Treatment", "Error", "Total"))

# display table
aov_tab

```

$\Rightarrow$ Environment is a factor affecting the yield of maize plants of 
family 1 (ANOVA, $F_{2, 147} = 205.7,$ 
$p = `r round(pf(205.7, 2, 147, lower.tail = FALSE), 3)`$).

## Task 3

```{r task_3_1}

exc_dat <- read.csv("data/Ex1.csv")

wh_mod <- lm(Weight ~ Height, data = exc_dat)

wh_mod_sum <- summary(wh_mod)
wh_mod_sum

coefs <- wh_mod_sum[["coefficients"]]

```

$\Rightarrow$ the value of the $t$-statistic for the `Intercept` is $-166$, the 
one of the regression coefficient for `Height` is $242.8$:

#### Intercept  
  
$T = \beta_{Intcp} / SE_{Intcp} =$
$`r round(coefs[1, 1], 2)` / `r round(coefs[1, 2], 2)` =$
$`r round(coefs[1, 1] / coefs[1, 2], 2)`$

#### Height  
  
$T = \beta_{Height} / SE_{Heigt} =$
$`r round(coefs[2, 1], 2)` / `r round(coefs[2, 2], 2)` =$
$`r round(coefs[2, 1] / coefs[2, 2], 2)`$
