---
title: "Exercise 1"
author: "Jonas Freimuth, 807218"
date: "27 10 2021"
output: pdf_document
---

```{r setup, include=FALSE}
require("ggplot2",     quietly = TRUE)
require("dplyr",       quietly = TRUE)
require("fastDummies", quietly = TRUE)
```

## Task 1  

Loading the data and checking dimensions:  

```{r task_1}
exc_dat <- read.csv("data/Ex1.csv")
str(exc_dat)
```

* `r nrow(exc_dat)` samples
* `r ncol(exc_dat)` variables, one discrete, two continuous

## Task 2

Histogram of body weights:

```{r task_2_1}
hist(exc_dat$Weight, xlab = "Weight", main = "Histogram of Weight",
     freq = FALSE)
lines(density(exc_dat$Weight), lwd = 2, col = "red")
```

The histogram suggests a bimodal distribution *possibly* due to the mixing of 
two normally distributed samples.

Histogram of body weights by gender:

```{r task_2_2}

plot(1, type = "n", xlab = "Weight", ylab = "Density",
     xlim = c(min(exc_dat$Weight), max(exc_dat$Weight)),
     ylim = c(0, 0.06),
     main = "Histogram of weight by gender",
     sub  = "Women: red; Men: blue")

hist(exc_dat$Weight[exc_dat$Gender == "Female"], freq = FALSE,
     col = "red", add = T)
lines(density(exc_dat$Weight[exc_dat$Gender == "Female"]), lwd = 2, col = "red")

hist(exc_dat$Weight[exc_dat$Gender == "Male"], freq = FALSE,
     col = "blue", add = T)
lines(density(exc_dat$Weight[exc_dat$Gender == "Male"]), lwd = 2, col = "blue")
```

Splitting by gender reveals two normally distributed samples.

## Task 3

```{r task_3_1}
plot(exc_dat$Height, exc_dat$Weight, 
     main = "Weight by height",
     xlab = "Height",
     ylab = "Weight")
text(x = mean(exc_dat$Height), y = max(exc_dat$Weight),
     labels = paste("r =",
                    round(cor(exc_dat$Weight, exc_dat$Height), 2)))

plot(exc_dat$Height, exc_dat$Weight,
     main = "Weight by height split by gender",
     sub  = "Women: black; Men: magenta",
     col  = as.factor(exc_dat$Gender),
     xlab = "Height",
     ylab = "Weight")
text(x = c(mean(exc_dat$Height[exc_dat$Gender == "Female"]),
           mean(exc_dat$Height[exc_dat$Gender == "Male"  ])),
     
     # set the women's r slightly lower to avoid overlap
     y = c(max(exc_dat$Weight) - 0.1 * max(exc_dat$Weight),
           max(exc_dat$Weight)),
     
     labels = c(paste("Women's r =",
                      round(cor(exc_dat$Weight[exc_dat$Gender == "Female"],
                                exc_dat$Height[exc_dat$Gender == "Female"]),
                            2)),
                paste("Men's r =",
                      round(cor(exc_dat$Weight[exc_dat$Gender == "Male"  ],
                                exc_dat$Height[exc_dat$Gender == "Male"  ]),
                            2))))
```

Correlations of weight with height become smaller when computing them for each 
gender alone as compared to computing them for the overall sample.

# Task 4

```{r task_4_1}
mod_1 <- lm(Weight ~ Height,          data = exc_dat)
mod_2 <- lm(Weight ~ Height + Gender, data = exc_dat)
mod_3 <- lm(Weight ~ Height * Gender, data = exc_dat)

summary(mod_1)
summary(mod_2)
summary(mod_3)
```
### Simple model:  

Coefficients: 
$\beta_0 = `r round(coef(mod_1)[1], 2)`$, 
$\beta_{Height} = `r round(coef(mod_1)[2], 2)`$.  

All parameters highly significant.

### Model with Gender: 

Coefficients: 
$\beta_0 = `r round(coef(mod_2)[1], 2)`$, 
$\beta_{Height} = `r round(coef(mod_2)[2], 2)`$,
$\beta_{Gender} = `r round(coef(mod_2)[3], 2)`$.

All parameters highly significant.

### Interaction Model: 

Coefficients: 
$\beta_0 = `r round(coef(mod_3)[1], 2)`$, 
$\beta_{Height} = `r round(coef(mod_3)[2], 2)`$,
$\beta_{Gender} = `r round(coef(mod_3)[3], 2)`$,
$\beta_{Height * Gender} = `r round(coef(mod_3)[3], 2)`$.

All parameters except the interaction term highly significant.

## Task 5

```{r task_5_1}
exc_dat_dummy <- dummy_cols(exc_dat, "Gender")

mod_dummy <- lm(Weight ~ Height + Gender_Male,
                data = exc_dat_dummy)

summary(mod_dummy)
```

When manually creating a dummy variable and fitting a model, the results are 
the same as when leaving the discrete variable.

```{r task_5_2}
exc_dat_recode <- exc_dat %>% 
  mutate(Gender = as.numeric(recode(Gender,
                                    !!! c("Male" = "5", "Female" = "1"))))

mod_recode <- lm(Weight ~ Height + Gender, data = exc_dat_recode)
summary(mod_recode)
```

However, when using arbitrary numeric values to represent discrete variable 
values, the differences between the numeric values have an effect. Here the 
parameter for gender becomes smaller.