---
title: "Exercise 10"
author: "Jonas Freimuth, 807218"
date: "19 01 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("lme4")
library("dplyr")
library("ggplot2")
library("data.table")
library("fastDummies")
```

## Task 1

```{bash task_1_1}
ls data/plink_data/exc_10/gwa
```

$\Rightarrow$ The file names indicate binary format.  

```{bash task_1_2}
plink --bfile data/plink_data/exc_10/gwa/gwa --noweb
```

$\Rightarrow$ 306102 SNPs in 4000 samples, 2000 males and females each, and 
2000 cases and control samples.

## Task 2

```{bash task_2_1}
mkdir data/plink_data/exc_10/gwa_thin

plink --bfile data/plink_data/exc_10/gwa/gwa --thin 0.1 --make-bed --out \
  data/plink_data/exc_10/gwa_thin/gwa_thin  --noweb
```

```{bash task_2_2}
mkdir -p data/plink_data/exc_10/out/all_test

plink --bfile data/plink_data/exc_10/gwa_thin/gwa_thin --maf 0.03 --geno 0.1 \
  --assoc --adjust --out data/plink_data/exc_10/out/all_test/all_test --noweb
```

```{r task_2_3}
assoc_res <- fread("data/plink_data/exc_10/out/all_test/all_test.assoc.adjusted")

head(assoc_res)

max_sig <- min(assoc_res$FDR_BY)
npv_sig <- sum(assoc_res$FDR_BY < 0.05)
```

$\Rightarrow$ After adjusting for multiple testing, the most significant 
association is at $p = `r max_sig`$, there are $`r npv_sig`$ SNPs with an 
interaction which is significant at all at $\alpha = 0.05$.

## Task 3

```{r task_3_1}
max_sig_adj <- max(assoc_res$UNADJ[assoc_res$FDR_BY < 0.05])

thr <- 3

n_bel_thr <- sum(-log10(assoc_res$UNADJ) > thr)

ggplot(assoc_res, aes(CHR, - log10(UNADJ), col = CHR)) +
  geom_jitter() + 
  geom_hline(yintercept = thr, lty = 2, col = "gray") +
  scale_color_viridis_c() + 
  labs(x = "Chromosome",
       y = "-log10(P)",
       title = "Manhattan plot",
       subtitle = paste0("Logistic regression,\nN(p < ", 10 ^ (- thr), ") = ",
                        n_bel_thr)) +
  theme_minimal() +
  theme(legend.position = "none")

```

## Task 4

```{bash task_4_1_1}
mkdir -p data/plink_data/exc_10/out/ca_test

plink --bfile data/plink_data/exc_10/gwa_thin/gwa_thin --maf 0.03 --geno 0.1 \
  --model --model-trend --adjust --out data/plink_data/exc_10/out/ca_test \
  --noweb
```

```{r task_4_1_2}
trend_res <- fread("data/plink_data/exc_10/out/ca_test.model.trend.adjusted")

head(trend_res)

max_sig <- min(trend_res$FDR_BY)
npv_sig <- sum(trend_res$FDR_BY < 0.05)

max_sig_adj <- max(trend_res$UNADJ[trend_res$FDR_BY < 0.05])

thr <- 3

n_bel_thr <- sum(-log10(trend_res$UNADJ) > thr)

ggplot(trend_res, aes(CHR, - log10(UNADJ), col = CHR)) +
  geom_jitter() + 
  geom_hline(yintercept = thr, lty = 2, col = "gray") +
  scale_color_viridis_c() + 
  labs(x = "Chromosome",
       y = "-log10(P)",
       title = "Manhattan plot",
       subtitle = paste0("Logistic regression,\nN(p < ", 10 ^ (- thr), ") = ",
                        n_bel_thr)) +
  theme_minimal() +
  theme(legend.position = "none")
```

$\Rightarrow$ After adjusting for multiple testing, the most significant 
association is at $p = `r max_sig`$, there are $`r npv_sig`$ SNPs with an 
interaction which is significant at all at $\alpha = 0.05$.

```{bash task_4_2_1}
mkdir -p data/plink_data/exc_10/out/log_reg

plink --bfile data/plink_data/exc_10/gwa_thin --maf 0.03 --geno 0.1 --logistic \
  --adjust --out data/plink_data/exc_10/out/log_reg --noweb
```

```{r task_4_2_2}
logistic_res <- fread("data/plink_data/exc_10/out/log_reg.assoc.logistic.adjusted")

head(logistic_res)

max_sig <- min(logistic_res$FDR_BY)
min_sig <- sum(logistic_res$FDR_BY < 0.05)

max_sig_adj <- max(logistic_res$UNADJ[logistic_res$FDR_BY < 0.05])

thr <- 3

n_bel_thr <- sum(-log10(logistic_res$UNADJ) > thr)

ggplot(logistic_res, aes(CHR, - log10(UNADJ), col = CHR)) +
  geom_jitter() + 
  geom_hline(yintercept = thr, lty = 2, col = "gray") +
  scale_color_viridis_c() + 
  labs(x = "Chromosome",
       y = "-log10(P)",
       title = "Manhattan plot",
       subtitle = paste0("Logistic regression,\nN(p < ", 10 ^ (- thr), ") = ",
                        n_bel_thr)) +
  theme_minimal() +
  theme(legend.position = "none")

```

$\Rightarrow$ After adjusting for multiple testing, the most significant 
association is at $p = `r max_sig`$, there are $`r npv_sig`$ SNPs with an 
interaction which is significant at all at $\alpha = 0.05$.


$\Rightarrow$ All methods give roughly the same results.

