---
title: "Exercise 2"
author: "Jonas Freimuth, 807218"
date: "3 11 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require("dplyr",      quietly = TRUE)
require("tidyr",      quietly = TRUE)
require("tidyselect", quietly = TRUE)
```

## Task 1

```{r task_1_1}
aov_tab <- data.frame("Sum Sq"  = c(720, 480, 1200),
                      "df"      = c(3, 16, 19),
                      "Mean Sq" = c(240, 30, NA),
                      "F val"   = c(8, NA, NA),
                      row.names = c("Treatment", "Error", "Total"))

aov_tab
```

Total degrees of freedom: $N - 1 = 19 \Rightarrow N = 20 \Rightarrow$ $20$ 
samples overall. Treatment degrees of freedom: $m - 1 = 3 \Rightarrow m = 4$ 
$\Rightarrow$ four treatment groups. Samples per group assuming perfect balance
of groups: $n = N / m = 20 / 4 = 5 \Rightarrow$ five samples per group.

There is a significant difference between the means of different treatments 
(ANOVA, $F_{3, 16} = 8, p = `r round(pf(8, 3, 16, lower.tail = FALSE), 3)`$).

## Task 2

```{r task_2_1}
mz_yld_raw <- read.csv("data/Ex2.csv")

str(mz_yld_raw)

mz_yld <- mz_yld_raw %>% 
  select(matches("E\\d_F\\d")) %>% 
  pivot_longer(cols      = everything(),
               names_to  = c("env", "fam"),
               names_sep = "_",
               values_to = "yield")
  

head(mz_yld)
```

```{r task_2_test}
mod_yld_f1 <- aov(yield ~ env, data = mz_yld %>% filter(fam == "F1"))

summary(mod_yld_f1)
```


```{r task_2_2}

aov_table <- function(y, x) {
  
  # ensure x and y have same length
  if (length(x) != length(y)) {
    stop("x and y must have the same lenth!")
  }
  
  # ensure x is a factor or character and y is numeric
  if (!(is.numeric(y))) {
    stop("y must be numeric!")
  }
  if (!(is.character(x) | is.factor(x))) {
    stop("x must be either a factor or a character vector!")
  }
  
  m <- length(unique(x))
  n <- length(x[x == unique(x[1])])
  N <- length(y)
  
  # calculate degrees of freedom
  dfs <- c("betw" = m - 1,
           "ins"  = N - m,
           "tot"  = N - 1)
  
  # calculate means
  grp_means <- data.frame(x = x,
                          y = y) %>% 
    group_by(x) %>% 
    summarise(mean = mean(y))
  
  gm <- mean(y)
  
  # calculate sums of squares
  ins_sum_sq_vec <- rep(NA, length(y))
  
  for (i in 1:m) {
    
    ins_sum_sq <- y[x == unique(x)[i]]
    ins_sum_sq <- (ins_sum_sq - grp_means$mean[i]) ^ 2
    ins_sum_sq_vec[(((i - 1) * n) + 1):((i * n))] <- ins_sum_sq
    
  }
  
  sum_sq <- c("betw" = sum((grp_means$mean  - gm) ^ 2 * n),
              "ins"  = sum(ins_sum_sq_vec),
              "tot"  = sum((y - gm) ^ 2))
  
  # calculate mean squares
  mean_sq <- c("betw" = sum_sq[["betw"]] / dfs[["betw"]],
               "ins"  = sum_sq[["ins" ]] / dfs[["ins" ]],
               "tot"  = NA)
  
  # calculate F value
  f_val <- c("betw" = mean_sq[["betw"]] / mean_sq[["ins"]],
             "ins"  = NA,
             "tot"  = NA)
  
  # assemble table
  aov_tab <- data.frame("Sum Sq"  = sum_sq,
                        "df"      = dfs,
                        "Mean Sq" = mean_sq,
                        "F val"   = f_val,
                        row.names = c("Treatment", "Error", "Total"))
  
  return(aov_tab)
  
}

mz_yld_f1 <- mz_yld %>% 
  filter(fam == "F1")

aov_tab <- aov_table(y = mz_yld_f1$yield, x = mz_yld_f1$env)

# display table
aov_tab

```

$\Rightarrow$ Environment is a factor affecting the yield of maize plants of 
family 1 (ANOVA, $F_{2, 147} = 205.7,$ 
$p = `r round(pf(205.7, 2, 147, lower.tail = FALSE), 3)`$).

## Task 3

```{r task_3_1}

exc_dat <- read.csv("data/Ex1.csv")

wh_mod <- lm(Weight ~ Height, data = exc_dat)

wh_mod_sum <- summary(wh_mod)
wh_mod_sum

coefs <- wh_mod_sum[["coefficients"]]

```

$\Rightarrow$ the value of the $t$-statistic for the `Intercept` is $-166$, the 
one of the regression coefficient for `Height` is $242.8$:

#### Intercept  
  
$T = \beta_{Intcp} / SE_{Intcp} =$
$`r round(coefs[1, 1], 2)` / `r round(coefs[1, 2], 2)` =$
$`r round(coefs[1, 1] / coefs[1, 2], 2)`$

#### Height  
  
$T = \beta_{Height} / SE_{Heigt} =$
$`r round(coefs[2, 1], 2)` / `r round(coefs[2, 2], 2)` =$
$`r round(coefs[2, 1] / coefs[2, 2], 2)`$
