---
title: "Exercise 3"
author: "Jonas Freimuth, 807218"
date: "8 12 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("EMSaov")
library("lme4")
```

## Task 1

```{r task_1_1}
snp_met_dat <- read.csv("data/Ex7_1.csv")

snp_met_aov <- anova(lm(metabolite ~ ., data = snp_met_dat))
snp_met_aov
```

```{r task_1_2}
row.names(snp_met_aov)[snp_met_aov[["Pr(>F)"]] <= 0.01]
```

```{r task_1_3}
for (snp in paste0("snp", c(1:8))) {
  cat("\n", snp, ":", "\n", sep = "")
  print(anova(lm(as.formula(paste("metabolite ~", snp)), snp_met_dat)))
}
```

## Task 2

```{r task_2_1}
tis_trs_dat <- read.csv("data/Ex7_2.csv")

t.test(transcription ~ tissue, tis_trs_dat)

anova(lm(transcription ~ tissue, tis_trs_dat))
```

$\Rightarrow$ Both $t$ and $F$ statistic will result in the same p-value in the
case of a two-mean comparison. Also the $F$-statistic is the square of the $t$-
statistic.

## Task 3

```{r task_3_1}
bw_dat <- read.csv("data/Ex7_3.csv")

bw_dat$age <- as.factor(bw_dat$age) 

anova(lm(weight ~ age * gender, bw_dat))
```

$\Rightarrow$ Means only differ between age groups, other factors have no 
influence.

```{r task_3_2}
EMSanova(weight ~ age + gender, bw_dat, type = c("R", "F"))
```
$\Rightarrow$ When modeling age as a random effect, we observe changes in 
significance due to changes in the way the F-values are computed: The effect of 
gender becomes even less significant.

## Task 4
```{r task_4_1}
pig_dat <- read.csv("data/Ex7_4.csv")

pig_lm <- lmer(production ~ sex + (1 | population), pig_dat)

summary(pig_lm)
```

```{r task_4_2}
ranef(pig_lm)$population
```

```{r task_4_3}
fixef_both <- fixef(pig_lm)

sex_lvls <- unique(pig_dat$sex)

fixef <- sex_lvls * fixef_both[["sex"]] + fixef_both[["(Intercept)"]]
names(fixef) <- sex_lvls

cat("Fixed effect coefficients for sex:", "\n")
fixef
```

