---
title: "Exercise 5"
author: "Jonas Freimuth, 807218"
date: "24 11 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("EMSaov")

# disable grouping output message
options(dplyr.summarise.inform = FALSE)

```

## Task 1

### Scenario (i)

$$ EMS_{A}   = \sigma^2_e + n \sigma_{\alpha \beta \gamma} + c n \sigma_{\alpha \beta}  + b n \sigma_{\alpha \gamma} + b c n \sigma^2_{\alpha}$$
$$EMS_{B}   = \sigma^2_e + n \sigma_{\alpha \beta \gamma} + a n \sigma_{\beta  \gamma} + c n \sigma_{\alpha \beta}  + a c n \sigma^2_{\beta}$$
$$EMS_{C}   = \sigma^2_e + n \sigma_{\alpha \beta \gamma} + b n \sigma_{\alpha \gamma} + a n \sigma_{\beta  \gamma} + a b n \sigma^2_{\gamma}$$
$$EMS_{AB}  = \sigma^2_e + n \sigma_{\alpha \beta \gamma} + c n \sigma_{\alpha \beta}$$
$$EMS_{AC}  = \sigma^2_e + n \sigma_{\alpha \beta \gamma} + b n \sigma_{\alpha \gamma}$$
$$EMS_{BC}  = \sigma^2_e + n \sigma_{\alpha \beta \gamma} + a n \sigma_{\beta  \gamma}$$
$$EMS_{ABC} = \sigma^2_e + n \sigma_{\alpha \beta \gamma}$$

### Scenario (ii)

$$EMS_{A}   = \sigma^2_e + n \sigma_{\alpha \beta \gamma} + c n \sigma_{\alpha \beta}  + b n \sigma_{\alpha \gamma} + b c n \theta^2_{\alpha}$$
$$EMS_{B}   = \sigma^2_e                                  + a n \sigma_{\beta  \gamma} + c n \sigma_{\alpha \beta}  + a c n \sigma^2_{\beta}$$
$$EMS_{C}   = \sigma^2_e                                  + b n \sigma_{\alpha \gamma} + a n \sigma_{\beta  \gamma} + a b n \sigma^2_{\gamma}$$
$$EMS_{AB}  = \sigma^2_e + n \sigma_{\alpha \beta \gamma} + c n \sigma_{\alpha \beta}$$
$$EMS_{AC}  = \sigma^2_e + n \sigma_{\alpha \beta \gamma} + b n \sigma_{\alpha \gamma}$$
$$EMS_{BC}  = \sigma^2_e                                  + a n \sigma_{\beta  \gamma}$$
$$EMS_{ABC} = \sigma^2_e + n \sigma_{\alpha \beta \gamma}$$


## Task 2

```{r task_2_1}
mz_yld_1 <- read.csv("data/Ex5_1.csv")

rnd_yld_aov <- EMSanova(yield ~ geno + env, mz_yld_1, type = c("R", "R"))
rnd_yld_aov
```

```{r task_2_2}
n <- 6
a <- 3
b <- 2

gen_MS <- rnd_yld_aov$MS[1]
env_MS <- rnd_yld_aov$MS[2]
int_MS <- rnd_yld_aov$MS[3]
err_MS <- rnd_yld_aov$MS[4]

gen_var_com <- err_MS + n * int_MS + b * n * gen_MS
env_var_com <- err_MS + n * int_MS + a * b * env_MS
int_var_com <- err_MS + n * int_MS
```

$\Rightarrow$ Variance components:  
* Genetic: $EMS_{gen} = `r gen_var_com`$  
* Environment: $EMS_{env} = `r env_var_com`$  
* Interaction: $EMS_{int} = `r int_var_com`$

The only component with significant variance at $\alpha = 0.01$ is the 
interaction of environment and genotype.

Each $F$-value is computed by the formula  
* $F_{A} = MS_A / MS_{AB}$, if A is a single effect  
* $F_{AB} = MS_{AB} / MS_E$, if AB is an interaction effect

## Task 3

```{r task_3_1}
mz_yld_2 <- read.csv("data/Ex5_2.csv")

nst_yld_aov <- EMSanova(yield ~ sire + dame, mz_yld_2, type = c("R", "R"),
                        nested = c(NA, "sire"))
nst_yld_aov
```


```{r task_3_2}
n <- 21
a <- 2
b <- 5

sir_MS <- nst_yld_aov$MS[1]
dam_MS <- nst_yld_aov$MS[2]
err_MS <- rnd_yld_aov$MS[3]

sir_var_com <- err_MS + n * dam_MS + b * n * sir_MS
dam_var_com <- err_MS + n * dam_MS
```


$\Rightarrow$ Variance components:  
* Sire: $EMS_{sir} = `r sir_var_com`$  
* Dame nested in sire: $EMS_{dam} = `r dam_var_com`$  

The only component with significant variance at $\alpha = 0.01$ is dame nested
in sire.  

Each $F$-value is computed by the formula:  
* $F_{A} = MS_A / MS_B$, if A is a single effect and B is nested in A   
* $F_{B} = MS_{B} / MS_E$, if B is a nested effect  
