---
title: "Exercise 2"
author: "Jonas Freimuth, 807218"
date: "3 11 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require("dplyr",      quietly = TRUE)
require("tidyr",      quietly = TRUE)
require("tidyselect", quietly = TRUE)
```

## Task 1

```{r task_1_1}
aov_tab <- data.frame("Sum Sq"  = c(720, 480, 1200),
                      "df"      = c(3, 16, 19),
                      "Mean Sq" = c(240, 30, NA),
                      "F val"   = c(8, NA, NA),
                      row.names = c("Treatment", "Error", "Total"))

aov_tab
```

Total degrees of freedom: $N - 1 = 19 \Rightarrow N = 20 \Rightarrow$ $20$ 
samples overall. Treatment degrees of freedom: $m - 1 = 3 \Rightarrow m = 4$ 
$\Rightarrow$ four treatment groups. Samples per group assuming perfect balance
of groups: $n = N / m = 20 / 4 = 5 \Rightarrow$ five samples per group.

There is a significant difference between the means of different treatments 
(ANOVA, $F_{3, 16} = 8, p = `r round(pf(8, 3, 16, lower.tail = FALSE), 3)`$).

## Task 2

```{r task_2_1}
mz_yld_raw <- read.csv("data/Ex2.csv")

str(mz_yld_raw)

mz_yld <- mz_yld_raw %>% 
  select(matches("E\\d_F\\d")) %>% 
  pivot_longer(cols      = everything(),
               names_to  = c("env", "fam"),
               names_sep = "_",
               values_to = "yield")
  

head(mz_yld)
```

```{r task_2_test}
mod_yld_f1 <- aov(yield ~ env, data = mz_yld %>% filter(fam == "F1"))

summary(mod_yld_f1)
```


```{r task_2_2}

aov_table <- function(y, x) {
  
  # ensure x and y have same length
  if (length(x) != length(y)) {
    stop("x and y must have the same lenth!")
  }
  
  # ensure x is a factor or character and y is numeric
  if (!(is.numeric(y))) {
    stop("y must be numeric!")
  }
  if (!(is.character(x) | is.factor(x))) {
    stop("x must be either a factor or a character vector!")
  }
  
  m <- length(unique(x))
  n <- length(x[x == unique(x[1])])
  N <- length(y)
  
  # calculate degrees of freedom
  dfs <- c("betw" = m - 1,
           "ins"  = N - m,
           "tot"  = N - 1)
  
  # calculate means
  grp_means <- data.frame(x = x,
                          y = y) %>% 
    group_by(x) %>% 
    summarise(mean = mean(y))
  
  gm <- mean(y)
  
  # calculate sums of squares
  ins_sum_sq_vec <- rep(NA, length(y))
  
  for (i in 1:m) {
    
    ins_sum_sq <- y[x == unique(x)[i]]
    ins_sum_sq <- (ins_sum_sq - grp_means$mean[i]) ^ 2
    ins_sum_sq_vec[(((i - 1) * n) + 1):((i * n))] <- ins_sum_sq
    
  }
  
  sum_sq <- c("betw" = sum((grp_means$mean  - gm) ^ 2 * n),
              "ins"  = sum(ins_sum_sq_vec),
              "tot"  = sum((y - gm) ^ 2))
  
  # calculate mean squares
  mean_sq <- c("betw" = sum_sq[["betw"]] / dfs[["betw"]],
               "ins"  = sum_sq[["ins" ]] / dfs[["ins" ]],
               "tot"  = NA)
  
  # calculate F value
  f_val <- c("betw" = mean_sq[["betw"]] / mean_sq[["ins"]],
             "ins"  = NA,
             "tot"  = NA)
  
  # assemble table
  aov_tab <- data.frame("Sum Sq"  = sum_sq,
                        "df"      = dfs,
                        "Mean Sq" = mean_sq,
                        "F val"   = f_val,
                        row.names = c("Treatment", "Error", "Total"))
  
  return(aov_tab)
  
}

mz_yld_f1 <- mz_yld %>% 
  filter(fam == "F1")

aov_tab <- aov_table(y = mz_yld_f1$yield, x = mz_yld_f1$env)

# display table
aov_tab

```

$\Rightarrow$ Environment is a factor affecting the yield of maize plants of 
family 1 (ANOVA, $F_{2, 147} = 205.7,$ 
$p = `r round(pf(205.7, 2, 147, lower.tail = FALSE), 3)`$).

## Task 3

```{r task_3_1}

exc_dat <- read.csv("data/Ex1.csv")

wh_mod <- lm(Weight ~ Height, data = exc_dat)

wh_mod_sum <- summary(wh_mod)
wh_mod_sum

coefs <- wh_mod_sum[["coefficients"]]

```

$\Rightarrow$ the value of the $t$-statistic for the `Intercept` is $-166$, the 
one of the regression coefficient for `Height` is $242.8$:

#### Intercept  
  
$T = \beta_{Intcp} / SE_{Intcp} =$
$`r round(coefs[1, 1], 2)` / `r round(coefs[1, 2], 2)` =$
$`r round(coefs[1, 1] / coefs[1, 2], 2)`$

#### Height  
  
$T = \beta_{Height} / SE_{Heigt} =$
$`r round(coefs[2, 1], 2)` / `r round(coefs[2, 2], 2)` =$
$`r round(coefs[2, 1] / coefs[2, 2], 2)`$

```{r task 3_2}

# define and calculate givens
pred     <- predict(wh_mod)
obs      <- exc_dat$Weight
mean_obs <- mean(obs)

n <- length(obs)
p <- 1

# calculate degrees of freedom
dfs <- c("reg" = p,
         "res" = n - p - 1,
         "tot" = n - 1)

# calculate sums of squares
sum_sq <- c("reg" = sum((pred - mean_obs) ^ 2),
            "res" = sum((obs  - pred    ) ^ 2),
            "tot" = sum((obs  - mean_obs) ^ 2))

# calculate mean squares
mean_sq <- c("reg" = sum_sq[["reg"]] / dfs[["reg"]],
             "res" = sum_sq[["res"]] / dfs[["res"]],
             "tot" = NA)

# calculate F value
f_val <- c("reg" = mean_sq[["reg"]] / mean_sq[["res"]],
           "res" = NA,
           "tot" = NA)

# assemble table
aov_tab <- data.frame("Sum Sq"  = sum_sq,
                      "df"      = dfs,
                      "Mean Sq" = mean_sq,
                      "F val"   = f_val,
                      row.names = c("Treatment", "Error", "Total"))

# display aov table
aov_tab

# calculate p value
pval <- pf(f_val[["reg"]], dfs[["reg"]], dfs[["res"]], lower.tail = FALSE)
```
*TODO: Explanation for DFs*
The degrees of freedom 

There is a significant influence of Height on Weight (ANOVA, 
$F_{1, 9998} = 58955.92, p = `r round(pval, 3)`$).

The coefficient of determination is: $R^2 = 1 - \frac{SS_{res}}{SS_{tot}}$
$= 1 - \frac{`r round(sum_sq[["res"]], 2)`}{`r round(sum_sq[["tot"]], 2)`}$
$= `r round(1 - sum_sq[["res"]] / sum_sq[["tot"]], 2)`$


## Task 4  

```{r task_4_1}
# Tukey HSD pos-hoc
mod_yld_f1_tukey <- TukeyHSD(mod_yld_f1)
mod_yld_f1_tukey

# pairwise t-tests for bonferroni and holm
pairs <- combn(unique(mz_yld_f1$env), 2)

pvals <- list()

for (i in 1:ncol(pairs)) {
  env_a <- mz_yld_f1$yield[mz_yld_f1$env == pairs[1, i]]
  env_b <- mz_yld_f1$yield[mz_yld_f1$env == pairs[2, i]]
  
  env_test <- t.test(env_a, env_b)
  
  pvals[[paste(pairs[1, i], pairs[2, i], sep = ", ")]] <- env_test$p.value
}

pvals <- unlist(pvals)

# correct pvals with bonferroni and holm method
pvals_bonf <- p.adjust(pvals, method = "bonferroni")
cat("\n\nBonferroni:\n"); pvals_bonf

pvals_holm <- p.adjust(pvals, method = "holm")
cat("\n\nHolm:\n"); pvals_holm


```

In contrast to the two methods based on simple t-tests with multiple testing
correction, Tukey's HSD method estimates the confidence intervals of the 
differences in a more appropriate way, and can therefore use the resulting 
p-values without multiple testing correction.

$\Rightarrow$ While the exact $p$-values returned by the different methods 
differ slightly, here the conclusions they suggest are equal.

## Task 5

```{r task_5_1}
# calculate and extract givens
a <- length(unique(mz_yld$env))
b <- length(unique(mz_yld$fam))

n <- nrow(mz_yld)
k <- n / (a * b)

y   <- mz_yld$yield
x_a <- mz_yld$env
x_b <- mz_yld$fam

# calculate means
grp_means_a <- mz_yld %>% 
  group_by(env) %>% 
  summarise(mean = mean(yield))


grp_means_b <-  mz_yld %>% 
  group_by(fam) %>% 
  summarise(mean = mean(yield))


grp_means_ab <- mz_yld %>% 
  group_by(env, fam) %>% 
  summarise(mean = mean(yield))
  

gm <- mean(y)

# calculate expected values for factor crossings
grp_expct_ab <- grp_means_ab %>% 
  select(env, fam) %>% 
  mutate(expct = NA)

grp_expct_ab <- apply(grp_expct_ab, 1, function(row) {
  mean_a <- grp_means_a$mean[grp_means_a$env == row[1]]
  mean_b <- grp_means_b$mean[grp_means_b$fam == row[2]]
  
  row[3] <- mean_a + mean_b - gm
  
  return(row)
}) %>% 
  t() %>% 
  as.data.frame() %>% 
  mutate(expct = as.numeric(expct))
  

# calculate degrees of freedom
dfs <- c("A"  =  a - 1,
         "B"  =  b - 1,
         "AB" = (a - 1) * (b  - 1),
         "E"  =  a * b  * (k  - 1),
         "T"  =  n - 1)

# calculate sums of squares within crossings of a and b
ins_ab_resid <- mz_yld %>% 
  group_by(env, fam) %>% 
  mutate(grp_resid = yield - mean(yield))

# compile sum of squares vector overall
sum_sq <- c("A"  = sum((grp_means_a $mean - gm)                 ^ 2) * b * k, 
            "B"  = sum((grp_means_b $mean - gm)                 ^ 2) * a * k,
            "AB" = sum((grp_means_ab$mean - grp_expct_ab$expct) ^ 2)     * k,
            "E"  = sum((ins_ab_resid$grp_resid)                 ^ 2), 
            "T"  = sum((y - gm)                                 ^ 2))

# calculate mean squares
mean_sq <- c("A"  = sum_sq[["A" ]] / dfs[["A" ]],
             "B"  = sum_sq[["B" ]] / dfs[["B" ]],
             "AB" = sum_sq[["AB"]] / dfs[["AB"]],
             "E"  = sum_sq[["E" ]] / dfs[["E" ]],
             "T"  = NA)

# calculate F value
f_val <- c("A"  = mean_sq[["A" ]] / mean_sq[["E"]],
           "B"  = mean_sq[["B" ]] / mean_sq[["E"]],
           "AB" = mean_sq[["AB"]] / mean_sq[["E"]],
           "E"  = NA,
           "T"  = NA)

# calcluate P vals
p_val <- c("A"  = pf(f_val[["A" ]], dfs[["A" ]], dfs[["E"]], lower.tail = FALSE),
           "B"  = pf(f_val[["B" ]], dfs[["B" ]], dfs[["E"]], lower.tail = FALSE),
           "AB" = pf(f_val[["AB"]], dfs[["AB"]], dfs[["E"]], lower.tail = FALSE),
           "E"  = NA,
           "T"  = NA)

# assemble table
aov_tab_int <- data.frame("Sum Sq"  = sum_sq,
                          "df"      = dfs,
                          "Mean Sq" = mean_sq,
                          "F val"   = f_val,
                          "P val"   = p_val,
                          row.names = c("Environment", "Family", "Interaction",
                                        "Error", "Total"))
aov_tab_int

```

$\Rightarrow$ This ANOVA-table shows a strongly significant effect of the 
interaction of Family with the environment, meaning some families produce larger
or smaller yield in certain environments than others. It also shows a strongly 
significant effect of both environment and family, which strongly suggests that 
both have an effect on the yield.

```{r task_5_2}

# calculate sums of squares within a
ins_ab_resid <- mz_yld %>% 
  group_by(env, fam) %>% 
  mutate(grp_resid = yield - mean(yield))
  

# calculate degrees of freedom
dfs <- c("A"  =  a - 1,
         "B"  =  b - 1,
         "E"  =  n - a - b + 1,
         "T"  =  n - 1)

# compile sum of squares vector overall
sum_sq <- c("A"  = sum((grp_means_a $mean - gm)                 ^ 2) * b * k, 
            "B"  = sum((grp_means_b $mean - gm)                 ^ 2) * a * k,
            "E"  = sum((ins_ab_resid$grp_resid)                 ^ 2) +
                   sum((grp_means_ab$mean - grp_expct_ab$expct) ^ 2)     * k,
            "T"  = sum((y - gm)                                 ^ 2))

# calculate mean squares
mean_sq <- c("A"  = sum_sq[["A" ]] / dfs[["A" ]],
             "B"  = sum_sq[["B" ]] / dfs[["B" ]],
             "E"  = sum_sq[["E" ]] / dfs[["E" ]],
             "T"  = NA)

# calculate F value
f_val <- c("A"  = mean_sq[["A" ]] / mean_sq[["E"]],
           "B"  = mean_sq[["B" ]] / mean_sq[["E"]],
           "E"  = NA,
           "T"  = NA)

# calcluate P vals
p_val <- c("A"  = pf(f_val[["A" ]], dfs[["A" ]], dfs[["E"]], lower.tail = FALSE),
           "B"  = pf(f_val[["B" ]], dfs[["B" ]], dfs[["E"]], lower.tail = FALSE),
           "E"  = NA,
           "T"  = NA)

# assemble table
aov_tab_smp <- data.frame("Sum Sq"  = sum_sq,
                          "df"      = dfs,
                          "Mean Sq" = mean_sq,
                          "F val"   = f_val,
                          "P val"   = p_val,
                          row.names = c("Environment", "Family",
                                        "Error", "Total"))
aov_tab_smp

```
$\Rightarrow$ This ANOVA-table shows shows a strongly significant effect of both
environment and family, strongly suggesting an effect of both on the yield, 
though the pvalues are not quite as strong as seen in the interaction table, and
no statement about whether some families produce greater or smaller yield in 
certain environments can be made.

